---
title: "STAT240 DIS343 Group 3 Project Draft"
output: html_document
date: "2026-04-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)

maternalmortraw<-read_csv("maternalmortraw.csv")

mort_chart = maternalmortraw %>% select(Location, Value, Period) %>% rename(Country = Location, Year = Period) %>% arrange(Country) %>% filter(Country %in% c("Nigeria", "India", "Poland", "Indonesia", "Russian Federation", "South Africa", "Venezuela (Bolivarian Republic of)", "United States of America"))

MaternalMortalityClean<-mort_chart %>% filter(row_number()%%2==!0)

antenatalraw<-read_csv("antenatalraw.csv")

AntenatalCareClean=antenatalraw %>% select(Location, Period, Value) %>% rename(Country = Location, Year = Period, propotionOfNeedsMet=Value) %>% arrange(Country) %>% filter(Country %in% c("Nigeria", "India", "Poland", "Indonesia", "Russian Federation", "South Africa", "Venezuela (Bolivarian Republic of)", "United States of America"))

MatMort<-c(155,226,1046,2.5,21,191,31,251) #got country average value by hand
antecov<-c(58.5,88.8,52.4,79.2,88,75.5,95.2,85.2)
Country<-c("India","Indonesia","Nigeria","Poland","Russia","South Africa", "USA", "Venezuela")
AnteMatMortTBL<-tibble(Country, antecov, MatMort)
AnteMatMortTBL
```

## Group Members

* Devika
* Quenby
* Roshan

## Introduction/Background
The availability of prenatal technology and care is crucial for the comfort and safety of expectant mothers. However, the gap in access and funding across different countries is continually widening. Before the deficit in some countries can be remedied, the actual needs and effects must be identified and studied. This inadequacy leads to the question, “How does lack of antenatal care impact maternal mortality and complication rates?” Based on our analysis, of the data, we hypothesize that an increase in antenatal care will lead to a decrease in maternal mortality. Many women around the world experience life threatening complications during pregnancy, birth, and after birth. These complications and causes of death include deaths from induced abortions without proper medical care, ectopic pregnancy, postpartum depression, hemorrhage, hypertension, sepsis, unsafe birthing conditions, obstructed labor, or other indirect causes (such as HIV/AIDS). The data represented may also be limited by the fact that much of this data’s collection happens at hospitals and other medical facilities. Therefore, pregnancy outcomes that do not occur at medical facilities and are not otherwise reported will not be included in general data. This limitation could artificially deflate certain countries' reported Maternal Mortality Rates. The complications associated with pregnancy and giving birth are often preventable. These preventative measures require funding and education for medical professionals and mothers which leads us to believe that people in countries with less access to prenatal care have higher mortality and complication rates. Maternal mortality is a public health issue that we believe increased access to antenatal care/resources would help to solve. 

## Background

### Data Sources
* Raw Data: The raw data we are using for this study is directly from the World Health Organization website. [WHO Maternal Mortality](https://www.who.int/data/gho/data/themes/topics/sdg-target-3-1-maternal-mortality) contains data on the maternal mortality by country. [WHO Maternal Mortality Ratio](hhttps://www.who.int/data/gho/data/indicators/indicator-details/GHO/maternal-mortality-ratio-(per-100-000-live-births)) contains data on the ratio of maternal mortality based on the ability to access pre-natal care.

### Variables
* Maternal Mortality: Maternal Mortality Ratio (deaths per 100,000 live births) for the year 2021 or the most recent checkpoint. This reflects the risk of maternal death in a single pregnancy, not cumulative lifetime risk.
* Antenatal Coverage:  Women who received antenatal care four or more times during pregnancy (%). A measure of access to and utilization of prenatal healthcare.

### Data Structure
We represented are choosing to represent our data with a linear regression graph. Our statistical tests allow us to observe the relationship between Maternal Mortality Rate and Antenatal Care Coverage over time in different countries. Looking at the slope of the regression from this test will allow us to test multiple factors in relation to Maternal Mortality Rate and will allow us to determine if this factor is impactful in predicting maternal mortality rate. Our linear regression model demonstrates a correlation between more Antenatal Care Coverage and lower maternal mortality rates.

### Data Cleaning and Processing
For the maternal mortality and antenatal care coverage data sets, the Location, Value, and Period columns were used, renaming them to Country and Year. The countries were also filtered in the sets such that they only included the eight that were being considered in the study. These new tables were organized by country and year, then using only the data for 2021 or the next most recent checkpoint. We created a linear regression analysis to compare the MMR with the ACC to determine if a correlation is present.


```{r echo = FALSE, warning = FALSE, message = FALSE}
ggplot(data=AnteMatMortTBL, aes(x=antecov, y=MatMort))+
  geom_point(mapping=aes(color=Country,size=1))+
  geom_smooth(method="lm", se=F)+
  labs(title = "Comparing the Percent of Women who recieved Antenatal Care Coverage to the Same Country's Maternal Mortality Rate", x="Women who Received Antenatal Care Four or More Times During Their Pregnancy(%)", y="Maternal Mortality Rate (Per 100,000 Live Births)")
```

Figure 1: A comparison of maternal mortality rate per 100,000 births to a percentage of women who received antenatal care four or more times during their pregnancy in eight select countries. Includes linear regression line in blue, indicating negative correlation.

```{r echo = FALSE, warning = FALSE, message = FALSE}
anteandmort<-data.frame(antecov,MatMort)
model2<-lm(data=anteandmort)
summary(model2)$coefficients
```


## Statistical Analysis
Our linear regression model demonstrates a correlation between more antenatal care coverage and lower maternal mortality rates. The multiple regression analysis yielded a p-value of .0563. This is marginally higher than the threshold of acceptable p-values, showing a weak, but likely present, correlation between the values. The slope of -.03 also suggests that the more ACC, the lower the MMRs will be.

### Parameter of Interest
The parameters of interest are the true mean maternal mortality rate, and its rate of change.

* $\beta_0$: The true mean maternal mortality rate for when the percentage of antenatal care access is 0%
* $\beta_1$: The change in mortality for every 1% increase in antenatal care access and coverage.

### Type of Inference
We will be doing a regression inference on the slope for our data.

### Assumptions
We assume that:

\[X = \text{Percentage of women who recieved antenatal care 4 or more times during their pregnancy}
\[Y = \text{Maternal mortality rate per 100,000 live births}
**Independance**: We assume that each country's data is independent of each other.

### Hypotheses
The hypotheses for our analysis are:

**Null Hypothesis (\(H_0\))**: There is no linear relationship between the percentage of women who recieved antenatal care 4 or more times during their pregnancy and the maternal mortality rate
**Alternative Hypothesis (\(H_{alpha}\))**: There is a negative linear relationship between the percentage of women recieving antenatal care 4 or more times and the maternal mortality rate

\[ H_0: \beta_x = \beta_y \implies \beta_x-\beta_y = 0 \\ H_{alpha}: \beta_x > \beta_y \implies \beta_x-\beta_y > 0 \]


### Test Statistic and Null Distribution

### Results

### Interpretation

### Confidence Interval

## Discussion

### Further Interpretation

Our analysis returns unclear results as to whether antenatal care coverage has a meaningful effect on maternal mortality rates between countries. The P-value we obtained is boarderline significant, and the presence of a correlation would align with previous global health research that has already documented that increased antenatal care is associated with a lower maternal mortality rate. 

### Shortcomings of the Analysis

Some possible limitations include confounding factors that are not represented in the data that we used to create the model, such as rural-urban disparities, ecological differences between countries, or a country's political or economic status, etc. There could also be bias based on the specific countries we selected and the limited number of locations represented in our study; however, there are often data quality problems, particularly related to the underreporting and misclassification of maternal deaths. 

### Additional Questions and Recommendations for Future Research

Therefore, data is often adjusted in order to take into account these data quality issues. Increasing the depth and volume of studies regarding the maternal mortality rate would be beneficial for future research on the topic and could help to save lives. 

### References